Extreme precip and heat waves: empirical-statistical downscaling and analysis

### 2016-09-27 CixPAG/CixPAG-imd.Rmd

---
title: "Dipper downscaling"
author: "Rasmus Benestad"
date: "August 22, 2016"
output: pdf_document
---

@@

## Example of downscaling a population of species

We can use the assumption of Poisson-type distribution for a count of e.g. population or events such as the bird "dipper" that lives in a colony in southern Norway. The dipper is a type of bird that lives in water falls. Its population is affected by the temperature and weather the water freezes during the winter.

This analysis was carried out for M. Gamelon, who provided the data on the dipper population

@@

```{r dipper}
## R-script to analyse dipper population

library(esd)
library(ncdf4)

examine <- FALSE

dipper <- read.table('~/data/dipper.csv',sep=',',header=TRUE)

dipper <- zoo(x=dipper[[2]],order.by=dipper[[1]])
y <- as.station(dipper,loc='Lyngdalselva',
                lon=7,lat=58.5,alt=520,
                param='population',unit='count',
                reference='Marlène Gamelon')

predictor <- retrieve("air.mon.mean.nc",lon=c(-1,17),lat=c(57,62))
## Only use the winter temperature:
predictor <- aggregate(subset(predictor,it='djf'),year,FUN='mean')

if (examine) {
  ## Check the distribution
  n <- seq(0,150,by=10); mu <- mean(dipper)
  hist(coredata(dipper),breaks=n,freq=FALSE,col='grey')
  lines(x,dpois(x,mu),lwd=2,col='red')

  ## Single ESD for inspection
  X <- EOF(predictor)
  z <- DS(y,X)
  plot(z,new=FALSE)
}
```

```{r}
## Downscale the dipper population directly based on the large-scale
## annual mean temperature

if (!file.exists('dipper.Z.rcp45.rda')) {
  Z.rcp45 <- DSensemble.annual(y,biascorrect=TRUE,
                               predictor=predictor,
                               lon=c(-1,17),lat=c(57,62),
                               abscoords=TRUE)
  save(file='dipper.Z.rcp45.rda',Z.rcp45)
} else load('dipper.Z.rcp45.rda')

if (!file.exists('dipper.Z.rcp85.rda')) {
  Z.rcp85 <- DSensemble.annual(y,biascorrect=TRUE,
                               rcp="rcp85",predictor=predictor,
                               lon=c(-1,17),lat=c(57,62),
                               abscoords=TRUE)
  save(file='dipper.Z.rcp85.rda',Z.rcp85)
} else load('dipper.Z.rcp85.rda')

if (!file.exists('dipper.Z.rcp26.rda')) {
  Z.rcp26 <- DSensemble.annual(y,biascorrect=TRUE,
                               rcp="rcp26",predictor=predictor,
                               lon=c(-1,17),lat=c(57,62),
                               abscoords=TRUE)
  save(file='dipper.Z.rcp26.rda',Z.rcp26)
} else load('dipper.Z.rcp26.rda')

year <- year(Z.rcp45)
ci90.rcp45 <- apply(coredata(Z.rcp45),1,quantile,
                    probs=c(0.05,0.95),na.rm=TRUE)
ci90.rcp26 <- apply(coredata(Z.rcp26),1,quantile,
                    probs=c(0.05,0.95),na.rm=TRUE)
ci90.rcp85 <- apply(coredata(Z.rcp85),1,quantile,
                    probs=c(0.05,0.95),na.rm=TRUE)

par(bty='n')
plot(range(year),range(ci90.rcp45,ci90.rcp26,ci90.rcp85,y),
     type='n',xlab='',ylab='Population',main='Dipper',
     sub=loc(y))
grid()
polygon(c(year,rev(year)),c(ci90.rcp85[1,],rev(ci90.rcp85[2,])),
                            col=rgb(1,0.5,0,0.3),border=rgb(1,0.5,0))
polygon(c(year,rev(year)),c(ci90.rcp45[1,],rev(ci90.rcp45[2,])),
                            col=rgb(0.5,1,0,0.3),border=rgb(0.5,1,0))
polygon(c(year,rev(year)),c(ci90.rcp26[1,],rev(ci90.rcp26[2,])),
                            col=rgb(0,0.5,1,0.3),border=rgb(0,0.5,1))
lines(year(y),coredata(y),lwd=4,pch=19,type='b')

legend(1900,200,c('RCP2.6','RCP4.5','RCP8.5'),
       col=c(rgb(0,0.5,1),rgb(0.5,1,0),rgb(1,0.5,0)),
       lwd=7,lty=1,bty='n')
legend(1905,170,'observed',lwd=4,pch=19,lty=1,bty='n')

```

>>>>

---
title: "CixPAG"
author: "Rasmus Benestad"
date: "August 3, 2016"
output: pdf_document
---

@@

@@

```{r esd}
library(esd)
```

## Reading data predictand

```{r pressure, echo=FALSE}
pt <- retrieve('data/IMD/tp_IMD.1951-2007.nc')
tx <- retrieve('data/IMD/maxT_IMD.1951-2013.nc')
map(pt)
```

### 2016-09-27 ClimaTrans/climatrans.Rmd

Model the number with heavy rainfall (more than 50 mm/day) in Delhi and Bangalore.

```{r fig.width=6, fig.height=6}
## Delhi
ndhr50.1.cal <- data.frame(y = coredata(subset(subset(ndhr50,is=1),it=c(1949,2004))),
                       X = coredata(subset(eof.es,it=c(1949,2004))))
ds.ndhr50.1 <- glm(y ~ X.1 + X.2 + X.3 + X.4 + X.5 + X.6 + X.7, data=ndhr50.1.cal,family="poisson")
plot(1949:2004,ndhr50.1.cal$y,pch=19,ylab=expression(N(X > 100*mm)),xlab='',main=loc(ndhr50)[1])
grid()
lines(1949:2004,exp(predict(ds.ndhr50.1)),col='red')

# Bangalore
ndhr50.3.cal <- data.frame(y = coredata(subset(subset(ndhr50,is=3),it=c(1949,2004))),
                       X = coredata(subset(eof.es,it=c(1949,2004))))
ds.ndhr50.3 <- glm(y ~ X.1 + X.2 + X.3 + X.4 + X.5 + X.6 + X.7, data=ndhr50.3.cal,family="poisson")
plot(1949:2004,ndhr50.3.cal$y,pch=19,ylab=expression(N(X > 100*mm)),xlab='',main=loc(ndhr50)[3])
grid()
lines(1949:2004,exp(predict(ds.ndhr50.3)),col='red')
```

@@

```{r}
print(paste('Mean number=',mean(ndhr100[,2],na.rm=TRUE),'Variance=',var(ndhr100[,2],na.rm=TRUE),
            'should equal',mean(ndhr100[,2],na.rm=TRUE),'for Poisson process'))
h <- hist(coredata(ndhr100[,2]),freq=FALSE,col='grey',
     main=loc(ndhr100)[2])
lines(h$mids-0.5,dpois(x=h$mids-0.5,lambda=mean(ndhr100[,2],na.rm=TRUE)),lwd=2,col='red')
grid()
```

@@

```{r}
print('Compare common EOFs')
diag.slp <- biasdiag(slp,it=month.abb[season],verbose=FALSE)
```

@@

The format of the station records bears a structure from the past where it was more important to take up little space rather than being human readable.

>>>>

Model the number with heavy rainfall (more than 50 mm/day) in Delhi and Bangalore. The results for Delhi and Bangalore also indicate a failure to capture the number of events.

```{r fig.width=6, fig.height=6}
## Delhi
ndhr50.1.cal <- data.frame(y = coredata(subset(subset(ndhr50,is=1),it=c(1949,2004))),
                       X = coredata(subset(eof.es,it=c(1949,2004))))
ds.ndhr50.1 <- glm(y ~ X.1 + X.2 + X.3 + X.4 + X.5 + X.6 + X.7, data=ndhr50.1.cal,family="poisson")
plot(1949:2004,ndhr50.1.cal$y,pch=19,ylab=expression(N(X > 50*mm)),xlab='',main=loc(ndhr50)[1])
grid()
lines(1949:2004,exp(predict(ds.ndhr50.1)),col='red')

# Bangalore
ndhr50.3.cal <- data.frame(y = coredata(subset(subset(ndhr50,is=3),it=c(1949,2004))),
                       X = coredata(subset(eof.es,it=c(1949,2004))))
ds.ndhr50.3 <- glm(y ~ X.1 + X.2 + X.3 + X.4 + X.5 + X.6 + X.7, data=ndhr50.3.cal,family="poisson")
plot(1949:2004,ndhr50.3.cal$y,pch=19,ylab=expression(N(X > 50*mm)),xlab='',main=loc(ndhr50)[3])
grid()
lines(1949:2004,exp(predict(ds.ndhr50.3)),col='red')
```

@@

```{r}
print(paste('Mean number=',round(mean(ndhr100[,2],na.rm=TRUE),1),
            'Variance=',round(var(ndhr100[,2],na.rm=TRUE),1),
            'should equal',round(mean(ndhr100[,2],na.rm=TRUE),1),'for Poisson process'))
h <- hist(coredata(ndhr100[,2]),freq=FALSE,col='grey',
     main=loc(ndhr100)[2])
lines(h$mids-0.5,dpois(x=h$mids-0.5,lambda=mean(ndhr100[,2],na.rm=TRUE)),lwd=2,col='red')
grid()
```

Also check whether the number of events in Delhi and Bangalore approximately follow a Poisson distribution.

```{r}
# Delhi
print(paste('Mean number=',round(mean(ndhr50[,1],na.rm=TRUE),1),
            'Variance=',round(var(ndhr50[,1],na.rm=TRUE),1),
            'should equal',round(mean(ndhr50[,1],na.rm=TRUE),1),'for Poisson process'))
h <- hist(coredata(ndhr50[,1]),freq=FALSE,col='grey',
     main=loc(ndhr50)[1])
lines(h$mids-0.5,dpois(x=h$mids-0.5,lambda=mean(ndhr50[,1],na.rm=TRUE)),lwd=2,col='red')
grid()

# Bangalore
print(paste('Mean number=',round(mean(ndhr50[,3],na.rm=TRUE),1),
            'Variance=',round(var(ndhr50[,3],na.rm=TRUE),1),
            'should equal',round(mean(ndhr50[,3],na.rm=TRUE),1),'for Poisson process'))
h <- hist(coredata(ndhr50[,3]),freq=FALSE,col='grey',
     main=loc(ndhr50)[3])
lines(h$mids-0.5,dpois(x=h$mids-0.5,lambda=mean(ndhr50[,3],na.rm=TRUE)),lwd=2,col='red')
grid()
```

The number of days with heavy rain is deviating somewhat from the Poisson distribution for the Indian data. We can compare the assumption of Poisson distribution with Norwegian data (all year) with known high quality (we have the metadata), and the results suggest a closer fit, although this data too is over-dispersed.

```{r}
data(bjornholt)
ndhr30 <- aggregate(bjornholt,year,'count',threshold=30)
print(paste('Mean number=',round(mean(ndhr30,na.rm=TRUE),1),'Variance=',round(var(ndhr30,na.rm=TRUE),1),
            'should equal',round(mean(ndhr30,na.rm=TRUE),1),'for Poisson process'))
h <- hist(coredata(ndhr30),freq=FALSE,col='grey',
     main=loc(ndhr30))
lines(h$mids-0.5,dpois(x=h$mids,lambda=mean(ndhr30,na.rm=TRUE)),lwd=2,col='red')
grid()
```

@@

```{r fig.width=12, fig.height=20}
print('Compare common EOFs')
diag.slp <- biasdiag(slp,it=month.abb[season],verbose=FALSE)
```

This test shows a comparison of the standard deviation of the principal components (PCs) associated with the reanalysis SLP and the GCM SLP over the sub-indian continent (region shown in the maps above) and the year-to-year variability in the monsoon season mean SLP. A good reproduction of the spatio-temporal variance structure for the SLP is expected to yield leading EOFs with similar variance and hence the blue symbols should be close to zero for $1 - \rho_{gcm}/\rho_{ref}$ along the x-axis. The results suggest otherwise, with large differences in the standard deviation connected to the PCs, and indicate that all GCMs have difficulties reproducing the monsoon features in the monsoon SLP.

@@

The format of the station records bears a structure from the past where it was more important to take up little space rather than being human readable. This cryptic and non-human readable way of recording data is associated several risks of error introduction, as it is easy to mistype some of the '0' entries and it is hard to spot obvious mistakes (e.g. unreasonable and unphysical values). Furthermore, reading data of this type of format also is prone to errors, as the data is stored in a way that may vary from time to time. There is also a high risk of the introduction of errors when reading such data.
